Channel Shaping Using Reconfigurable Intelligent Surfaces: From Diagonal to Beyond

### 2024-01-08 main.tex
:pencil: Update literature review

% glossaries-extra
\glsdisablehyper
\setabbreviationstyle[acronym]{long-short}
\newacronym{ao}{AO}{Alternating Optimization}
\newacronym{bd}{BD}{Beyond-Diagonal}
\newacronym{bcd}{BCD}{Block Coordinate Descent}
\newacronym{dof}{DoF}{Degree of Freedom}
\newacronym{mimo}{MIMO}{Multiple-Input Multiple-Output}
\newacronym{rcg}{RCG}{Riemannian Conjugate Gradient}
\newacronym{ris}{RIS}{Reconfigurable Intelligent Surface}
\newacronym{pc}{PC}{Point-to-point Channel}
\newacronym{ic}{IC}{Interference Channel}
\newacronym{snr}{SNR}{Signal-to-Noise Ratio}
\newacronym{wsr}{WSR}{Weighted Sum-Rate}
\newacronym{mmse}{MMSE}{Minimum Mean-Square Error}
\newacronym{wmmse}{WMMSE}{Weighted \gls{mmse}}
\newacronym{mse}{MSE}{Mean-Square Error}

@@

\begin{abstract}
	\gls{ris} is a promising technology that recycles and redistributes ambient waves for improved wireless performance.
	% \gls{ris} is a real-time programmable reflector that recycles and redistributes ambient waves for improved wireless performance.
	This paper investigates the use of passive \gls{ris} for channel shaping in \gls{mimo} \gls{pc} and \gls{ic}.
	We depart from the widely-adapted diagonal phase shift model and focus on a general asymmetric \gls{bd} architecture, which allows off-diagonal entries by in-group connections between elements.
	For arbitrary group size, we propose a closed-form iterative algorithm for quadratic problems and a \gls{rcg} algorithm for general problems.
	Various case studies are then conducted, including rate maximization and Pareto singular value characterization for \gls{pc}, and interference alignment and \gls{wsr} maximization for \gls{ic}.
	To quantify channel shaping capability, we also derive analytical bounds under specific scenarios and validate those by aforementioned methods.
	Simulation results suggest the advantage of \gls{bd} \gls{ris} scales with group size, while dyadic or tetradic architecture usually strikes a good balance between performance and complexity.
	% Through theoretical analysis and numerical evaluation,
	Our findings unlock unprecedented efficiency and adaptability, paving the way for smarter and greener wireless environments.
	% truly programmable
	% useful guidelines for prototype design.
\end{abstract}

@@

\begin{section}{Introduction}
	The quest for reliable, high-speed, and ubiquitous wireless connectivity has been long-standing since Marconi's illuminating radio in 1895.
	Great successes have been made at transmitter and receiver sides over the past century, and the society is unprecedentedly close to the Shannon limit \cite{Shannon1948}.
	Today we are witnessing a paradigm shift from \emph{connectivity} to \emph{intelligence}, where the wireless environment is no longer a chaotic medium but a conscious agent that serves on demand.
	This is made possible by the recent advances in \gls{ris}, a real-time programmable metasurface made of low-power reflecting elements.
	It can manipulate the amplitude, phase, frequency, and polarization of the scattered waves \cite{Basar2019} with a higher energy efficiency, lower cost, lighter footprint, and greater scalability than conventional relays.
	Using \gls{ris} for \emph{passive beamforming} has attracted significant interest in wireless communication, backscatter, sensing, and power transfer literature \cite{Wu2019,Wu2020c,Yang2020,Zheng2021,Jia2020,Liang2022,Liu2022a,Hua2023,Wu2021d,Feng2022,Zhao2022}, reporting a second-order array gain and fourth-order power scaling law (with proper waveform).
	On the other hand, \gls{ris} also enables \emph{backscatter modulation} by dynamically switching between different patterns, as investigated \cite{Liu2019d,Karasik2020,Basar2020,Ma2020a,Zhang2021d,Hua2022,Zhao2022a} and prototyped \cite{Tang2019a,Dai2020a}.
	Although fruitful outcomes have been harvested by fancy optimization tools, one fundamental yet critical unanswered question is the \emph{channel shaping} capability of \gls{ris}: To what extent can \gls{ris} reshape the wireless channel?

	% we aim to answer this question by a simple yet insightful approach: \emph{from diagonal to beyond}.
	% The answer depends heavily on the phase shift model of \gls{ris}.
	Before answering the question, we first need to revisit the \gls{ris} architecture.

	% Despite the fruitful results, the existing literature on \gls{ris} is still limited to the diagonal phase shift model, where the phase shifters are constrained to be diagonal matrices.

>>>>

% glossaries-extra
\glsdisablehyper
\setabbreviationstyle[acronym]{long-short}
\newacronym{ao}{AO}{Alternating Optimization}
\newacronym{bd}{BD}{Beyond-Diagonal}
\newacronym{bcd}{BCD}{Block Coordinate Descent}
\newacronym{dof}{DoF}{Degree of Freedom}
\newacronym{siso}{SISO}{Single-Input Single-Output}
\newacronym{miso}{MISO}{Multiple-Input Single-Output}
\newacronym{mimo}{MIMO}{Multiple-Input Multiple-Output}
\newacronym{rcg}{RCG}{Riemannian Conjugate Gradient}
\newacronym{ris}{RIS}{Reconfigurable Intelligent Surface}
\newacronym{pc}{PC}{Point-to-point Channel}
\newacronym{ic}{IC}{Interference Channel}
\newacronym{snr}{SNR}{Signal-to-Noise Ratio}
\newacronym{wsr}{WSR}{Weighted Sum-Rate}
\newacronym{mmse}{MMSE}{Minimum Mean-Square Error}
\newacronym{wmmse}{WMMSE}{Weighted \gls{mmse}}
\newacronym{mse}{MSE}{Mean-Square Error}

@@

\begin{abstract}
	\gls{ris} recycles and redistributes ambient waves for improved wireless performance.
	% \gls{ris} is a promising technology that recycles and redistributes ambient waves for improved wireless performance.
	% \gls{ris} is a real-time programmable reflector that recycles and redistributes ambient waves for improved wireless performance.
	This paper investigates the use of passive \gls{ris} for channel shaping in \gls{mimo} \gls{pc} and \gls{ic}.
	We depart from the widely-adapted diagonal phase shift model and focus on a general asymmetric \gls{bd} architecture, which allows off-diagonal entries by in-group connections between elements.
	For arbitrary group size, we propose a closed-form iterative algorithm for quadratic problems and a \gls{rcg} algorithm for general problems.
	Various case studies are then conducted, including Pareto singular value characterization and rate maximization for \gls{pc}, and interference alignment and \gls{wsr} maximization for \gls{ic}.
	To quantify channel shaping capability, we also derive analytical bounds under specific scenarios and validate those by aforementioned methods.
	Simulation results suggest the advantage of \gls{bd} \gls{ris} scales with group size, while dyadic or tetradic architecture usually strikes a good balance between performance and complexity.
	% Through theoretical analysis and numerical evaluation,
	Our findings unlock unprecedented efficiency and adaptability, paving the way for smarter and greener wireless environments.
	% truly programmable
	% useful guidelines for prototype design.
\end{abstract}

@@

\begin{section}{Introduction}
	% The quest for reliable, high-speed, and ubiquitous wireless connectivity has been long-standing since Marconi's illuminating radio in 1895.
	% Great successes have been made at transmitter and receiver sides over the past century, and the society is unprecedentedly close to the Shannon limit \cite{Shannon1948}.
	Today we are witnessing a paradigm shift from {connectivity} to {intelligence}, where the wireless environment is no longer a chaotic medium but a conscious agent that serves on demand.
	This is made possible by the recent advances in \gls{ris}, a real-time programmable metasurface made of numerous non-resonant sub-wavelength scattering elements.
	It can manipulate the amplitude, phase, frequency, and polarization of the scattered waves \cite{Basar2019} with a higher energy efficiency, lower cost, lighter footprint, and greater scalability than conventional relays.
	Using \gls{ris} for \emph{passive beamforming} has attracted significant interest in wireless communication \cite{Wu2019,Wu2020c,Yang2020,Zheng2021}, backscatter \cite{Jia2020,Liang2022}, sensing \cite{Liu2022a,Hua2023}, and power transfer literature \cite{Wu2021d,Feng2022,Zhao2022}, reporting a second-order array gain and fourth-order power scaling law (with proper waveform).
	On the other hand, \gls{ris} also enables \emph{backscatter modulation} by dynamically switching between different patterns, as already investigated \cite{Karasik2020,Basar2020,Zhao2022a} and prototyped \cite{Tang2019a,Dai2020a}.
	Although fruitful outcomes have been harvested over optimization tools, one fundamental yet critical question is the \emph{channel shaping} capability: To what extent can a passive \gls{ris} reshape the wireless channel?
	The answer indeed depends on the hardware architecture and scattering model.

	In conventional (a.k.a. diagonal) \gls{ris}, each scattering element is tuned by a dedicated impedance and functions as an \emph{individual} phase shifter \cite{Wu2020}.
	% , accounting for a diagonal phase shift model with unit magnitude.
	The idea is extended to \gls{bd} \gls{ris} \cite{Shen2020a,Li2023b} featuring in-group connections between elements.
	This allows \emph{cooperative} reflection --- wave impinging on one element can propagate within the circuit and depart partially from all elements.
	\gls{bd} \gls{ris} can thus control both amplitude and phase of the reflected wave, generalizing the scattering matrix from diagonal with unit-magnitude entries to block diagonal with  unitary blocks.
	Its benefit has been recently shown in receive power maximization \cite{Nerini2023,Santamaria2023,Fang2023,Nerini2023a}, transmit power minimization \cite{Zhou2023}, and rate maximization \cite{Zhou2023,Nerini2023a,Li2023d,Li2023c,Bartoli2023}.
	The channel estimation \cite{Li2023e} and mutual coupling \cite{Li2023f} issues have also been investigated.
	Therefore, \gls{bd} \gls{ris} is envisioned as the next generation channel shaper with stronger signal processing flexibility \cite{Li2023g}.

	% Attempts to characterize the channel shaping capability can be classified into \emph{singular value centric} and \emph{power centric} methods.
	Metrics of channel shaping can be classified into \emph{singular value centric} and \emph{power centric}.
	For diagonal \gls{ris}, the former has been studied in terms of minimum singular value \cite{ElMossallamy2021}, effective rank \cite{ElMossallamy2021,Meng2023}, condition number \cite{Zheng2022,Huang2023} in \gls{pc}, and degree of freedom \cite{Bafghi2022,Zheng2023,Chae2023} in \gls{ic}.
	The latter has been studied in terms of channel power \cite{Wu2019} in \gls{pc} and leakage interference \cite{Santamaria2023a} in \gls{ic}.
	When it comes to \gls{bd} \gls{ris}, the only metric analyzed is the channel power \cite{Nerini2023,Santamaria2023} where the \gls{mimo} case is still unexplored.\footnote{Single-stream \gls{mimo} with given precoder and combiner was considered in \cite{Nerini2023}. It is equivalent \gls{siso} channel shaping.}
	Although insightful, the answers are far from complete and the optimization techniques are limited to very specific scenarios.
	In this paper, we aim to quantify the channel shaping capability of \gls{bd} \gls{ris} through theoretical analysis and numerical evaluation.
	The contributions are summarized below.

	% we aim to answer this question by a simple yet insightful approach: \emph{from diagonal to beyond}.
	% The answer depends heavily on the phase shift model of \gls{ris}.
